Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-01-30 SCP2008/paper/metamonads.tex
deal with a few more comments

Therefore, in MetaOCaml, code must be generated just right (see
\cite{TahaThesis} for many simple examples).  For more complex
examples, new techniques are necessary, e.g., abstract interpretation
\cite{KiselyovTaha}.  But more problems remain
\cite{scp-CohenDGHKP06}: generating binding forms (``names'')
when generating loop bodies or conditional branches, and making
continuation-passing style (CPS) code clear.  Many authors
understandably shy away from CPS code as it quickly becomes
unreadable.  But this is needed for proper name generation.
The problems of compositionality of code generators, expressing
dependencies among them and domain-specific knowledge remain.

@@

We reuse the English word ``aspect'' for the various facets of the family
of LU decomposition algorithms.  While our use shares the 
spirit of aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
our implementation methodology is radically different\footnote{However it seems
that we are closer to the original ideas of AOP~\cite{709568,mendhekar97rg}
which were also concerned with scientific software}.  We firmly believe that
our typed generative methodology is much better suited to functional
programming, compared to attempts to graft the program-trace-based
methodology of object-oriented versions of AOP.

@@

To see the benefit of code generation, we notice that we can easily
parameterize our code:

@@

let retN (a : ('c,'v)code) : 
 (<classif: 'c; answer: ('c,'w)code; ..>,('c,'v)code) monad 
   = fun s k -> .<let t = .~a in .~(k s .<t>.)>.

@@

In Figure~\ref{ourmonad}, the monad (yielding values of type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and a value, and
yields an answer of type |w|.  The monad is polymorphic over the
three type parameters, which would require |monad| to be a type
constructor with three arguments. When we use this monad for code
generation, we will need yet another type variable, environment
classifier \cite{taha-environment}. With type constructors taking more
and more arguments, it becomes increasingly difficult to read and write
types -- which we will be doing extensively when writing module
signatures in Section~\ref{functors}. The fact that OCaml renames all type
variables when printing out types confuses matters further. An elegant
solution to these sorts of problems has been suggested by 
Jacques Garrigue on the Caml mailing list.
%% (cited from
%% \url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
%% no date, no exact ref). 
%
We use a single type parameter |'p| to
represent all parameters of our monad (all parameters but the type of
the monadic value |'v|). The type variable |'p| is constrained to be
the type of an object with methods (fields) |state| and |answer|. The
object may include more fields, represented by |..|. Values of that
type are not part of our computations and need not exist. We merely
use the object type, as an convenient way to specify extensible
\emph{type-level} records in OCaml.

@@

The operation |retN a| is the let-insertion operation, whose simpler
version we called |letgen| earlier. It is the first computation with
a control effect: indeed, the result of |retN a| is \emph{not} the
result of invoking its continuation |k|. Rather, its result is a |let|
code expression. Such a behavior is symptomatic of control operators
(in particular, |abort|).

@@

This encapsulation of staging annotations into code combinators lets
us easily change the output form of our generated LU programs.  For example,
We may use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
values, for benchmarking and regression tests purposes. We could also use
code combinators producing C or Fortran code.

>>>>

Therefore, in MetaOCaml, code must be generated just right (see
\cite{TahaThesis} for many simple examples).  For more complex
examples, new techniques are necessary, e.g., abstract interpretation
\cite{KiselyovTaha}.  But more problems remain
\cite{scp-CohenDGHKP06}: generating binding forms (``names'')
when generating loop bodies or conditional branches, and making
continuation-passing style (CPS) code clear.  Many authors
understandably shy away from CPS code as it quickly becomes
unreadable.  But this is needed for proper name generation.
To be able to build modular code generators, three important problems
remain: compositionality, expressing dependencies, and integration of
domain-specific knowledge.

@@

We reuse the English word ``aspect'' for the various facets of the family
of LU decomposition algorithms.  While our use shares the 
spirit of aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
our implementation methodology is radically different\footnote{However it seems
that we are closer to the original ideas of AOP~\cite{709568,mendhekar97rg}
which were also concerned with scientific software.}.  We firmly believe that
our typed generative methodology is much better suited to functional
programming, compared to attempts to graft the program-trace-based
methodology of object-oriented versions of AOP.

@@

By parametrizing our code, we can make the benefits of code generation
evident:

@@

let retN (a : ('c,'v) code) : 
 (<classif: 'c; answer: ('c,'w) code; ..>,('c,'v) code) monad 
   = fun s k -> .<let t = .~a in .~(k s .<t>.)>.

@@

In Figure~\ref{ourmonad}, the monad (yielding values of type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and a value, and
yields an answer of type |w|.  The monad is polymorphic over the
three type parameters, which would require |monad| to be a type
constructor with three arguments. When we use this monad for code
generation, we will need yet another type variable for the environment
classifiers \cite{taha-environment}. With type constructors taking more
and more arguments, it becomes increasingly difficult to read and write
types -- which we will be doing extensively when writing module
signatures in Section~\ref{functors}. The fact that OCaml renames all type
variables when printing out types confuses matters further. An elegant
solution to these sorts of problems has been suggested by 
Jacques Garrigue on the Caml mailing list.
%% (cited from
%% \url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
%% no date, no exact ref). 
%
We use a single type parameter |'p| to
represent all parameters of our monad (all parameters but the type of
the monadic value |'v|). The type variable |'p| is constrained to be
the type of an object with methods (fields) |state| and |answer|. The
object may include more fields, represented by |..|. Values of that
type are not part of our computations and need not exist. We merely
use the object type, as an convenient way to specify extensible
\emph{type-level} records in OCaml.

@@

The operation |retN a| is the let-insertion operation, whose simpler
version we called |letgen| earlier. It is the first computation with
a control effect: indeed, the result of |retN a| is \emph{not} the
result of invoking its continuation |k|. Rather, its result is a |let|
code expression. Such behavior is symptomatic of control operators
(in particular, |abort|).  The name can be taken to mean \emph{return
a Named computation}; the name allows for proper sharing, but otherwise 
|retN| is used in writing generators in the same way as |ret|.

@@

This encapsulation of staging annotations into code combinators lets
us easily change the output form of our generated LU programs.  For example,
we may use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
values, for benchmarking and regression tests purposes. We could also use
code combinators producing C or Fortran code.
